Integrated multi-omics analysis reveals Lactobacillus anti-inflammatory process in vaginal tissue

### 2021-02-15 reports/rmarkdown/03_integrated_datasets.Rmd
fixes plots and adds picrust

```{r, echo = FALSE}
PATH <- "~/Desktop/Desktop_stuff/NBIS/SMS_Projects/broliden_5325/" #Path to the data

@@

  pvalues <- lapply(enrichments,function(x){
    temp <- allpvals
    temp[x$pathway] <- x$pval
    return(temp)
  })
  pvalues <- t(as.data.frame(pvalues))
  pvalues <- -log10( pvalues )
  pvalues[is.na(pvalues)] <- 1

  allNES <- setNames(rep(NA,length(allpathways)),allpathways)
  NESes <- lapply(enrichments,function(x) {
    temp <- allNES
    temp[x$pathway] <- x$NES
    return(temp)
  })
  NESes <- t(as.data.frame(NESes))
  NESes[is.na(NESes)] <- 0
  NESes[ pvalues < -log10(0.1) ] <- 0

@@

@@

>>>>

```{r, echo = FALSE}
PATH <- "../../" #Path to the data

@@

  pvalues <- lapply(enrichments,function(x){
    temp <- allpvals
    temp[x$pathway] <- x$pval
    return(temp)
  })
  pvalues <- t(as.data.frame(pvalues))
  pvalues <- -log10( pvalues )
  pvalues[is.na(pvalues)] <- 0

  allNES <- setNames(rep(NA,length(allpathways)),allpathways)
  NESes <- lapply(enrichments,function(x) {
    temp <- allNES
    temp[x$pathway] <- x$NES
    return(temp)
  })
  NESes <- t(as.data.frame(NESes))
  # NESes[is.na(NESes)] <- 0
  # NESes[ pvalues < -log10(0.1) ] <- 0

@@

write.csv2(NESse_list,"../../results/Hallmark_NESse_list.csv")

@@

write.csv2(NESse_list,"../../results/Kegg_NESse_list.csv")

### 2021-02-15 reports/rmarkdown/manuscript/manuscript_template.Rmd
fixes plots and adds picrust

# I usually load my libraries up front to keep things organized
library(bookdown)
library(knitr)
library(kableExtra)
library(ggplot2)
library(igraph)
library(dplyr)
library(stringr)
library(rafalib)
# remotes::install_github("czarnewski/niceRplots")
library(niceRplots)

@@

```{r, echo = FALSE, eval = TRUE}
# Load reasult from the analysis
metadata <- read.csv2("../../../results/metadata_integration.csv",row.names = 1)
patient_groups <- metadata$joint_clustering
bac_communities <- as.matrix(read.csv2("../../../results/bacterial_communities.csv",row.names = 1))[,1]
SNN_patients <- read.csv("../../../results/patient_SNN_graph.csv",row.names = 1)
SNN_bacteria <- read.csv("../../../results/bacteria_SNN_graph.csv",row.names = 1)
clean_Ftest <- read.csv2("../../../results/metadata_association_results.csv",row.names = 1)
datasets_all_samples <- readRDS("../../../results/datasets_all_samples.RDS")

@@

Patients were thus subdivided into `r length(unique(patient_groups))` groups,

@@

par(mar=c(0,0,0,0))
plot( g , vertex.label.cex=0.000001 , vertex.color = pal[factor(patient_groups)] ,
      edge.width=  ( E(g)$weight / max(E(g)$weight)) ,
      vertex.size=10,main="", 
      edge.color=colorRampPalette(c("grey95","black"))(90) [ round( E(g)$weight / max(E(g)$weight) * 88 )+1 ] )
legend("topright",title.adj = 0,
       legend = levels(factor(patient_groups)),xjust = 0,yjust = 1,
       bty = "n",pch = 21,pt.bg = pal,pt.cex = 1,xpd=T)

@@

barlist( data = rbind(  BV=mL ,
                        counts_16S=xL),
         main = "Lactobacillus",
         genes = c("BV","counts_16S"),
         clustering = patient_groups,
         draw_mean_lines=F,col = pal)

@@

barlist( data = rbind(  BV=mM ,
                        counts_16S=xM),
         main = "Monbilicus",
         genes = c("BV","counts_16S"),
         clustering = patient_groups,
         draw_mean_lines=F,col = pal)

barlist( data = rbind(  BV=mG ,
                        counts_16S=xG),
         main = "Gardanerella",
         genes = c("BV","counts_16S"),
         clustering = patient_groups,
         draw_mean_lines=F,col = pal)

@@

shann <- vegan::diversity(t(x),index = c("shannon") )
simp <- vegan::diversity(t(x),index = c("simpson") )
invsimp <- vegan::diversity(t(x),index = c("invsimpson") )
divers <- as.matrix(rbind(shann,simp,invsimp) )
divers[is.infinite(divers)] <- 0
rownames(divers) <- c("Shannon","simpson","Inv Simpson")
barlist( data = divers,
         main = "",
         genes = c("Shannon","simpson","Inv Simpson"),
         clustering = patient_groups,
         draw_mean_lines=F,col = pal)

@@

###################################
# BACTERIAL VAGINOSIS COMPARISSON #
###################################
par(mar=c(4,5,1,12))
temp <- t(t(2^x-1)/colSums(2^x-1))
taxa <- read.csv2("../../../results/taxonomy.csv",row.names = 1,stringsAsFactors = F,as.is = T)
taxa[is.na(taxa)] <- "NA"
taxa <- taxa[rownames(temp),]
temp <- rowsum(temp , taxa$Class)
temp[is.nan(temp)] <- 0
o <- order( rowSums(temp) , decreasing = T)
sample_o <- order( patient_groups )
temp <- temp[o,sample_o]*100
temp <- temp[-grep("NA",rownames(temp)),]
temp <- rbind(temp[1:16,], Other = colSums(temp[17:nrow(temp),]) )

@@

@@

@@

@@

@@

@@

>>>>

# I usually load my libraries up front to keep things organized
library(bookdown)
library(knitr)
library(kableExtra)
library(ggplot2)
library(igraph)
library(dplyr)
library(stringr)
library(rafalib)
library(scales)
# remotes::install_github("czarnewski/niceRplots",force=T)
library(niceRplots)

@@

```{r, echo = FALSE, eval = TRUE}
# Load reasult from the analysis
metadata <- read.csv2("../../../results/metadata_integration.csv",row.names = 1)
patient_groups <- factor(setNames(metadata$joint_clustering,metadata$ID))
bac_communities <- as.matrix(read.csv2("../../../results/bacterial_communities.csv",row.names = 1))[,1]

SNN_patients <- read.csv("../../../results/patient_SNN_graph.csv",row.names = 1)
SNN_bacteria <- read.csv("../../../results/bacteria_SNN_graph.csv",row.names = 1)
clean_Ftest <- read.csv2("../../../results/metadata_association_results.csv",row.names = 1)
datasets_all_samples <- readRDS("../../../results/datasets_all_samples.RDS")

picrust_data <- read.table("../../../data/pred_metagenome_unstrat.tsv",header = T,row.names = 1)
picrust_dge <- read.csv("../../../data/Picrust2_top_functions.csv",header = T,row.names = 1)
```

```{r, echo = FALSE, eval = TRUE}
#Assign names to patient groups
big_groups <- c(
  L.inners=    "5",
  L.inners2=   "4",
  L.crispatus= "2",
  Gard=        "3",
  Gard.Prev=   "6",
  Gard.Prev2=  "8",
  Gard.Lacto=  "7",
  Patho=       "1"
)

group_annotation <- as.character(patient_groups)
group_annotation <- names(big_groups)[match(group_annotation,big_groups)]
group_annotation <- as.factor(setNames(group_annotation,names(patient_groups)))

metadata$group_annotation <- group_annotation
```

```{r, echo = FALSE, eval = TRUE}
bac_communities <- factor(paste0("BC",sprintf("%02d",bac_communities)))
bacs <- names(bac_communities)

#Assign names to bacterial communities
BC_groups <- c(
  BC01= "BC01_Bifidobacterium",
  BC02= "BC02_Mycoplasma",
  BC03= "BC03_Lactobacillus.spp",
  BC04= "BC04_Escherichia.sp",
  BC05= "BC05_Other1",
  BC06= "BC06_Other2",
  BC07= "BC07_Corynebacterium.spp",
  BC08= "BC08_Other3",
  BC09= "BC09_Ureaplasma.sp",
  BC10= "BC10_Campylobacter",
  BC11= "BC11_Mobiluncus_BVAB",
  BC12= "BC12_Faecalibacterium",
  BC13= "BC13_Dialister",
  BC14= "BC14_P.timonesis",
  BC15= "BC15_T.scotoductus"
)

bac_communities <- as.character(bac_communities)
bac_communities <- BC_groups[match(bac_communities,names(BC_groups))]
bac_communities <- as.factor(setNames(bac_communities,bacs))
```

@@

Patients were thus subdivided into `r length(unique(group_annotation))` groups,

@@

par(mar=c(0,0,0,5))
plot( g , vertex.label.cex=0.000001 , vertex.color = pal[factor(group_annotation)] ,
      edge.width=  ( E(g)$weight / max(E(g)$weight)) ,
      vertex.size=10,main="", 
      edge.color=colorRampPalette(c("grey95","black"))(90) [ round( E(g)$weight / max(E(g)$weight) * 88 )+1 ] )
legend("topright",title.adj = 0,
       legend = levels(factor(group_annotation)),xjust = 0,yjust = 1,
       bty = "n",pch = 21,pt.bg = pal,pt.cex = 1,xpd=T)

@@

barlist( data = rbind(  BV=mL ,
                        counts_16S=xL),
         main = "Lactobacillus",
         genes = c("BV","counts_16S"),
         clustering = group_annotation,
         draw_mean_lines=F,col = pal)

@@

barlist( data = rbind(  BV=mM ,
                        counts_16S=xM),
         main = "Monbilicus",
         genes = c("BV","counts_16S"),
         clustering = group_annotation,
         draw_mean_lines=F,col = pal)

barlist( data = rbind(  BV=mG ,
                        counts_16S=xG),
         main = "Gardanerella",
         genes = c("BV","counts_16S"),
         clustering = group_annotation,
         draw_mean_lines=F,col = pal)

@@

shann <- vegan::diversity(t(x),index = c("shannon") )
simp <- vegan::diversity(t(x),index = c("simpson") )
invsimp <- vegan::diversity(t(x),index = c("invsimpson") )
divers <- as.matrix(rbind(shann,simp,invsimp) )
divers[is.infinite(divers)] <- 0
rownames(divers) <- c("Shannon","simpson","Inv Simpson")
barlist( data = divers,
         main = "",
         genes = c("Shannon","simpson","Inv Simpson"),
         clustering = group_annotation,
         draw_mean_lines=F,col = pal)

@@

#######################
# BACTERIAL ABUNDANCE #
#######################
par(mar=c(4,5,1,12))
temp <- t(t(2^x-1)/colSums(2^x-1))
taxa <- read.csv2("../../../results/taxonomy.csv",row.names = 1,stringsAsFactors = F,as.is = T)
taxa[is.na(taxa)] <- "NA"
taxa <- taxa[rownames(temp),]
temp <- rowsum(temp , taxa$Class)
temp[is.nan(temp)] <- 0
o <- order( rowSums(temp) , decreasing = T)
sample_o <- order( group_annotation )
temp <- temp[o,sample_o]*100
temp <- temp[-grep("NA",rownames(temp)),]
temp <- rbind(temp[1:16,], Other = colSums(temp[17:nrow(temp),]) )

@@

```{r, echo = FALSE, eval = TRUE}
# Define Layout and set labels
layout(matrix(c(2 ,3,4,5,      6,7,17,17,          19,19,19,19,
                1,1,1,1,      1,8,17,17,          20,20,20,20,
                1,1,1,1,      1,9,18,18,         21,21,24,24,
                1,1,1,1,      1,10,18,18,         22,22,24,24,
                16,15,14,13,  12,11,23,23,         23,23,24,24 ) , ncol= 12,byrow = T),
       widths = 1,heights = c(1.2,1.2,1.2,1.2,1.2))
# layout.show(12)
figlabels <- letters

#########################
# BACTERIAL COMMUNITIES #
#########################
gB <- graph_from_adjacency_matrix(as.matrix(SNN_bacteria), mode = "undirected",diag = F,weighted = T)
set.seed(1)
lB <- layout_with_fr(gB,niter=3000,start.temp=30)

par(mar=c(0,0,0,5))
plot( gB , vertex.label.cex=0.000001 , vertex.color = pal[factor(bac_communities)] ,
      edge.width=  ( E(gB)$weight / max(E(gB)$weight)) ,
      vertex.size=10,main="", 
      edge.color=colorRampPalette(c("grey95","black"))(90) [ round( E(gB)$weight / max(E(gB)$weight) * 88 )+1 ] )
legend("topright",title.adj = 0,
       legend = levels(factor(bac_communities)),xjust = 0,yjust = 1,
       bty = "n",pch = 21,pt.bg = pal,pt.cex = 1,xpd=T)

par(mar=c(1,1,1,1))
for(j in sort(unique(bac_communities)) ){
  empty_plot()
  text( 0,1,labels = paste0(rownames(SNN_bacteria)[ bac_communities == j ],collapse = "\n"),adj=c(0,1),cex=.5)
  title(main=paste0("community",j),cex.main=.7)
}
for(i in 1:(18-length(unique(bac_communities)))){empty_plot()}

###################################
# BACTERIAL VAGINOSIS COMPARISSON #
###################################
taxa <- read.csv2("../../../results/taxonomy.csv",row.names = 1,stringsAsFactors = F,as.is = T)
taxa[is.na(taxa)] <- "NA"

par(mar=c(1,7,2,3))
x <- taxa[match(rownames(SNN_bacteria),rownames(taxa)),2]
y <- taxa[match(rownames(SNN_bacteria),rownames(taxa)),3]
# plot_sankey( data.frame( x[y!="NA"] , y[y!="NA"]), color_by = 1,
#        plot_labels = T, plot_weights = F,order_1_by = "total",
#        order_2_by = "total",use_w1 = T,use_w2 = T,gapv = .02,gap2v = 0,smoothing = .1,
#        pal=hue_pal()(length(unique(x))))

# plot_sankey( data.frame( y[y!="NA"] , bac_communities[y!="NA"]), color_by = 1,
#      plot_labels = T, plot_weights = F,order_1_by = "total",
#      order_2_by = "total",use_w1 = T,use_w2 = T,gapv = .02,gap2v = 0,smoothing = .1,
#      pal=hue_pal()(length(unique(y))))

plot_sankey( data.frame( x[y!="NA"] , bac_communities[y!="NA"]), color_by = 1,
     plot_labels = T, plot_weights = F,order_1_by = "total",
     order_2_by = "total",use_w1 = T,use_w2 = T,gapv = .02,gap2v = 0,smoothing = .1,
     pal=hue_pal()(length(unique(x))))

add_letter(figlabels[1]); figlabels <- figlabels[-1]

##############################################
# BACTERIAL COMMUNITY DIFFERENTIAL ABUNDANCE #
##############################################
par(mar=c(1,7,1,3))

module_means_per_dataset <- lapply(datasets_all_samples[-1],function(x){
  return( rowsum( x[rownames(SNN_bacteria),] , group = bac_communities ) / c(table(bac_communities)) )
})

for( j in c("ASV_tissue_V3_normalized_batch_corrected",
        "ASV_CVL_V3_normalized_batch_corrected",
        "ASV_CVL_V2_normalized_NOT_batch_corrected")){
  temp_group <- as.character(group_annotation)
  temp_group[ is.na(temp_group) ] <- "NA"
  temp_group[ temp_group == "" ] <- "NA"
  temp_group <- factor(temp_group)

  barlist(data=module_means_per_dataset[[j]],
          genes=rownames(module_means_per_dataset[[j]]),col = pal,
          clustering=temp_group, main=paste0("\n", sub("_norm.*","",j)))
}
add_letter(figlabels[1]); figlabels <- figlabels[-1]

empty_plot()

#####################
# BACTERIAL PICRUST #
#####################
par(mar=c(1,7,1,3))
# 
# library(keggorthology)
# data("KOgraph")
# annot <- as.data.frame(unlist(nodeData(KOgraph,attr="tag")))
# colnames(annot)[1] <- "KEGGID"
# annot$pathway <- rownames(annot)
# annot$KEGGID <- paste0("K",annot$KEGGID)
# 
# library(KEGGREST)

download.file("https://raw.githubusercontent.com/picrust/picrust2/master/picrust2/default_files/description_mapfiles/ko_info.tsv.gz","../../../data/KEGG_pathways_to_KO.tsv")
download.file("https://raw.githubusercontent.com/picrust/picrust2/master/picrust2/default_files/pathway_mapfiles/KEGG_pathways_to_KO.tsv","../../../data/KEGG_pathways_to_KO.tsv")
download.file("https://raw.githubusercontent.com/picrust/picrust2/master/picrust2/default_files/description_mapfiles/KEGG_pathways_info.tsv.gz","../../../data/KEGG_pathways_info.tsv.gz")
system("gunzip ../../../data/*.gz")

kegg_pathway_annot <- fgsea::gmtPathways('../../../data/KEGG_pathways_to_KO.tsv')

kegg_names <- read.delim('../../../data/KEGG_pathways_info.tsv',header = F)

ko_names <- readLines('../../../data/ko_info.tsv')
ko_names <- strsplit(ko_names,"\t")

path_trend <- t(sapply(kegg_pathway_annot,function(x){
  x <- rownames(picrust_data)[rownames(picrust_data) %in% x]
  if(length(x)>=2){
    return(colSums(picrust_data [ x , ]))
  } else {
    return(rep(0, ncol(picrust_data)))
  }
}))

path_trend <- path_trend[rowSums(path_trend)!=0 , ]
rownames(path_trend) <- kegg_names[ match(rownames(path_trend),kegg_names[,1]),2]

library(edgeR)
design <- model.matrix(~  group_annotation, data=metadata)

y <- DGEList(counts=floor(path_trend), remove.zeros = T)
y <- calcNormFactors(y,method = "TMM")
y <- estimateGLMCommonDisp(y,design)
y <- estimateGLMTagwiseDisp(y,design)
fit <- glmFit(y, design)

lrt <- glmLRT(fit,coef=1:8)
top <- topTags(lrt,adjust.method = "BH",n = "all",sort.by = "p.value")[[1]]
colnames(top) <- sub("group_annotation","",colnames(top))
top <- cbind(logFC.Gard=0,top)

top_dge <- (top$PValue < 0.01) #& (rowSums( abs(top[,grep("logFC",colnames(top))]) >= log2(1.5) ) >= 1)
top_dge <- rownames(top)[ top_dge ]

norm_path_trend <- log2( t( t(path_trend) / colSums(path_trend) ) * 1000 + 1 )
norm_path_trend <- norm_path_trend[top_dge, ]
norm_path_trend <- t(apply(norm_path_trend,1,function(x){scale(x,T,T)}))
norm_path_trend[norm_path_trend > 5] <- 5
colnames(norm_path_trend) <- colnames(path_trend)
sample_use <- colnames(norm_path_trend)

mypar(2,4,mar=c(1,.2,2,.2))
h <- hclust( as.dist( (1- cor(t(norm_path_trend)))/2 ), method = "ward.D2")
plot( rev(as.dendrogram(h)) , horiz = T, leaflab = "none",xaxs="i",yaxs="i",axes=F)
abline(v=1,xpd=F,col="red",lty=2)
gene_module <- cutree(h, h = 1)

points( rep(0,length(gene_module)),
        seq(length(gene_module),1,length.out = length(gene_module)),
        col=taxa_pal[factor(gene_module[h$order])],
        pch=16,cex=.5)

image( t(norm_path_trend[h$order,][nrow(norm_path_trend):1,order(group_annotation[sample_use])]),
       col = colorRampPalette(c("navy","navy","grey95","firebrick4","firebrick4") )(99),
       breaks = seq(-5,5,length.out = 100),axes=F)
points( seq(0,1,length.out = length(sample_use)),rep(par("usr")[4],length(sample_use)),
        col=pal[group_annotation[sample_use][ order(group_annotation[sample_use])]],
        pch=16)

par(mar=c(4,4,2,2))
set.seed(1)
UMAP_picrust <- uwot::umap(t(norm_path_trend),n_neighbors = 20,
                 metric = "correlation",min_dist = .2,spread = .4,
                 repulsion_strength = .4,negative_sample_rate = 3)
plot(UMAP_picrust,bg=pal[group_annotation[sample_use]],main="UMAP on top functional profile",
     pch=21,xlab="UMAP1",ylab="UMAP2",frame=F,axes=F)
```

@@

TRX <- datasets_all_samples[["Tissue_RNAseq_V3_normalized"]]
TRX_counts <- round(2^TRX - 1)
TRX_counts <- TRX_counts [ rowSums(TRX_counts>0)>= 1 , ]
sample_use <- colnames(TRX_counts)[colSums(TRX_counts)!=0]
TRX_counts <- TRX_counts[,sample_use]

library(edgeR)

design <- model.matrix(~  group_annotation, data=metadata[sample_use,])

y <- DGEList(counts=TRX_counts[,sample_use], remove.zeros = T)
y <- calcNormFactors(y,method = "TMM")
y <- estimateGLMCommonDisp(y,design)
y <- estimateGLMTagwiseDisp(y,design)
fit <- glmFit(y, design)

lrt <- glmLRT(fit,coef=2:8)
top <- topTags(lrt,adjust.method = "BH",n = "all",sort.by = "p.value")[[1]]
colnames(top) <- sub("group_annotation","",colnames(top))
top <- cbind(logFC.Gard=0,top)

top_dge <- (top$PValue < 0.01) #& (rowSums( abs(top[,grep("logFC",colnames(top))]) >= log2(1.5) ) >= 1)
top_dge <- rownames(top)[ top_dge ]

top_TRX_counts <- cpm(y)[top_dge, ]
top_TRX_counts <- t(apply(top_TRX_counts,1,function(x){scale(x,T,T)}))
top_TRX_counts[top_TRX_counts > 5] <- 5
colnames(top_TRX_counts) <- colnames(TRX_counts)

mypar(2,4,mar=c(1,.2,2,.2))
h <- hclust( as.dist( (1- cor(t(top_TRX_counts)))/2 ), method = "ward.D2")
plot( rev(as.dendrogram(h)) , horiz = T, leaflab = "none",xaxs="i",yaxs="i",axes=F)
abline(v=2,xpd=F,col="red",lty=2)
gene_module <- cutree(h, h = 2)

points( rep(0,length(gene_module)),
        seq(length(gene_module),1,length.out = length(gene_module)),
        col=taxa_pal[factor(gene_module[h$order])],
        pch=16,cex=.5)

image( t(top_TRX_counts[h$order,][nrow(top_TRX_counts):1,order(group_annotation[sample_use])]),
       col = colorRampPalette(c("navy","navy","grey95","firebrick4","firebrick4") )(99),
       breaks = seq(-5,5,length.out = 100),axes=F)
points( seq(0,1,length.out = length(sample_use)),rep(par("usr")[4],length(sample_use)),
        col=pal[group_annotation[sample_use][ order(group_annotation[sample_use])]],
        pch=16)

set.seed(1)
UMAP_TRX <- uwot::umap(t(top_TRX_counts),n_neighbors = 10,
                 metric = "correlation",min_dist = 0.1,spread = 5,
                 negative_sample_rate = 10)
plot(UMAP_TRX,bg=pal[group_annotation[sample_use]],main="UMAP on top DGE RNAseq",
     pch=21,xlab="UMAP1",ylab="UMAP2",frame=F,axes=F)
```

@@

```{r, echo = FALSE, eval = TRUE, fig.asp=1}
figlabels <- letters
temp_legend <- NULL

layout(matrix(c(1,7,
                2,7,
                3,7,
                4,8,
                5,8,
                6,8) , ncol= 2,byrow = T),widths = c(4,1),heights = c(.5,1,.5,1,.5,1))

for(i in c("ASV_CVL_V3_normalized_batch_corrected",
           "ASV_CVL_V2_normalized_NOT_batch_corrected",
           "ASV_tissue_V3_normalized_batch_corrected")){
  x <- datasets_all_samples[[i]]

  par(mar=c(1,5,1,1))
  temp <- t(t(2^x-1)/colSums(2^x-1))

  shann <- vegan::diversity(t(temp),index = c("shannon") )
  simp <- vegan::diversity(t(temp),index = c("simpson") )
  invsimp <- vegan::diversity(t(temp),index = c("invsimpson") )
  divers <- as.matrix(rbind(shann,simp,invsimp) )
  divers[is.infinite(divers)] <- 0
  rownames(divers) <- c("Shannon","simpson","Inv Simpson")
  barlist( data = divers[,order(group_annotation)],
         main = "",
         genes = c("Shannon","simpson","Inv Simpson"),
         # clustering = group_annotation,
         draw_mean_lines=F,col = pal[factor(group_annotation[order(group_annotation)])])

  par(mar=c(3,5,1,1))
  # taxa <- read.csv2("../../../results/taxonomy.csv",row.names = 1,stringsAsFactors = F,as.is = T)
  # taxa[is.na(taxa)] <- "NA"
  # taxa <- taxa[rownames(temp),]
  # temp <- rowsum(temp , taxa$Class)
  temp[is.nan(temp)] <- 0

  if(is.null(temp_legend)){ 
    temp_legend <- temp   #Use the same legend from the 1st dataset
  }

  o <- order( rowSums(temp_legend) , decreasing = T)
  sample_o <- names(group_annotation)[ order( group_annotation )]
  temp <- temp[o,sample_o]*100
  # temp <- temp[-grep("NA",rownames(temp)),]
  temp <- rbind(temp[1:16,], Other = colSums(temp[17:nrow(temp),]) )

  taxa_pal <- c(RColorBrewer::brewer.pal(8,"Pastel2"),RColorBrewer::brewer.pal(8,"Pastel1"),"grey90")
  barplot( temp, las=2 , border = NA , yaxs="i", xaxs="i",ylab="abundance",main=sub("_nor.*","",i),font.main=1,
           col = taxa_pal[factor(rownames(temp),levels = rownames(temp))] )
  # legend(x = par("usr")[c(2)],
  #        y = par("usr")[c(4)],title.adj = 0,
  #        legend = rownames(temp),xjust = 0,yjust = 1,
  #        bty = "n",pch = 22,pt.cex = 2,cex = 1,xpd=T,
  #        pt.bg = taxa_pal)
}

par(mar=c(4,0,1,0))
empty_plot()
legend(x = par("usr")[c(1)],
         y = par("usr")[c(4)],title.adj = 0,
         legend = levels(group_annotation),xjust = 0,yjust = 1,
         bty = "n",pch = 22,pt.cex = 2,cex = 1,xpd=T,title = "patient groups",
         pt.bg = pal)

empty_plot()
legend(x = par("usr")[c(1)],
         y = par("usr")[c(4)],title.adj = 0,
         legend = rownames(temp),xjust = 0,yjust = 1,
         bty = "n",pch = 22,pt.cex = 2,cex = 1,xpd=T,
          title="top abundant bacteria",
         pt.bg = taxa_pal)

@@

```{r, echo = FALSE, eval = TRUE, fig.asp=.5}
figlabels <- letters
temp_legend <- NULL

mypar(1,4,mar=c(12,1,1,0))

for(i in c("ASV_CVL_V3_normalized_batch_corrected",
           "ASV_CVL_V2_normalized_NOT_batch_corrected",
           "ASV_tissue_V3_normalized_batch_corrected")){

  continuous_parameters <- c("age","Count_CD4_v2","Count_CD4_v3","Plasma_S_Prog_nmol_L_v3","Plasma_S_Estradiol_pmol_L_v3","Conc_T4_v2","Conc_T3_v2","Conc_Progesteron_v2","Conc_Estradiol_v2","Conc_Cortisol_v2","Freq_CD4_v2","Freq_CD4_v3","Freq_Lympho_v2","Freq_Lympho_v3","visit_clients7days_v2")
  continuous_metadata <- metadata[,continuous_parameters]
  # continuous_metadata <- cbind(continuous_metadata , t(module_means_per_dataset[[i]]) )

  metadata_cor <- cor(t(module_means_per_dataset[[i]]),continuous_metadata,use = "pairwise.complete.obs")
  # metadata_cor_test <- cor.test(t(module_means_per_dataset[[i]]),continuous_metadata)

  image( t(metadata_cor [nrow(metadata_cor):1,]),axes=F,
         breaks = seq(-1,1,length.out = 90),main=sub("_nor.*","",i),font.main=1,
         col=colorRampPalette(c("navy","grey95","firebrick"))(89) )
  # text(  par("usr")[2]+.02 ,seq(1,0,length.out = nrow(metadata_cor)) , 
  #        labels = rownames(metadata_cor), srt = 0, adj = c(0,.5), xpd = TRUE, cex=1)
  text( seq(0,1,length.out = ncol(metadata_cor)) , par("usr")[3]-.02 , 
        labels = colnames(metadata_cor), srt = 90, adj = c(1,.5), xpd = TRUE, cex=1)

  #add label
  add_letter(figlabels[1]); figlabels <- figlabels[-1]
}

par(mar=c(12,0,1,0))
image( t(metadata_cor [nrow(metadata_cor):1,]),axes=F,
         breaks = seq(-1,1,length.out = 90),main="",
         col=colorRampPalette(c("white"))(89) )
text(  0 ,seq(1,0,length.out = nrow(metadata_cor)) , 
  labels = rownames(metadata_cor), srt = 0, adj = c(0,.5), xpd = TRUE, cex=1)
```

@@

```{r, echo = FALSE, eval = TRUE}
figlabels <- letters
mypar(1,2,mar=c(3,3,3,10))

datasets <- datasets_all_samples[grep("ASV",names(datasets_all_samples))]
datasets <- lapply(datasets,t)
datasets <- t( as.data.frame(datasets) )

group_means <- sapply(levels(group_annotation), function(x){
  rowMeans( datasets[ , group_annotation == x ] )
})

d <- as.dist( (1 - cor(group_means)) / 2 )
h <- hclust( d , method = "ward.D2")
plot( as.dendrogram(h) , horiz=T, xlab="height")
points( rep(0 , length(h$labels)) , 1:length(h$labels) , pch=21, cex=1.5,
        bg=pal[factor(h$labels[ h$order ],levels = h$labels)] )
add_letter(figlabels[1]); figlabels <- figlabels[-1]

datasets <- datasets_all_samples[grep("ASV",names(datasets_all_samples))]
datasets <- as.data.frame(datasets)

group_means <- sapply(levels(factor(bac_communities)), function(x){
  colMeans( datasets[ bac_communities == x , ] )
})

d <- as.dist( (1 - cor(group_means)) / 2 )
h <- hclust( d , method = "ward.D2")
plot( as.dendrogram(h) , horiz=T, xlab="height")
points( rep(0 , length(h$labels)) , 1:length(h$labels) , pch=21, cex=1.5,
        bg=taxa_pal[factor(h$labels[ h$order ],levels = h$labels)] )
add_letter(figlabels[1]); figlabels <- figlabels[-1]
```

@@

```{r, echo = FALSE, eval = TRUE,fig.asp=.3}
figlabels <- letters
for(j in names(module_means_per_dataset)){
  mypar(2,8,mar=c(0.5,0.5,4,0.5))

  plot(UMAP_picrust,bg=pal[group_annotation],main="patient groups",
     pch=21,xlab="UMAP1",ylab="UMAP2",frame=F,axes=F)
  add_letter(figlabels[1]); figlabels <- figlabels[-1]

  for(i in 1:nrow(module_means_per_dataset[[j]])){

    xx <- module_means_per_dataset[[j]][i,]
    xx <- xx / max(xx)
    o <- order(xx)
    plot(UMAP_picrust[o,],
         col=colorRampPalette( c("grey95","navy")) (99)[ 1 + (xx[o]*98) ],
         main=rownames(module_means_per_dataset[[j]])[i],
         pch=16,xlab="UMAP1",ylab="UMAP2",frame=F,axes=F,font.main=1,cex.main=1)
  }
}
```

## Figure S5

```{r, echo = FALSE, eval = TRUE,fig.asp=.3}
figlabels <- letters
for(j in names(module_means_per_dataset)){
  mypar(2,8,mar=c(0.5,0.5,4,0.5))

  plot(UMAP_TRX,bg=pal[group_annotation[sample_use]],main="patient groups",
     pch=21,xlab="UMAP1",ylab="UMAP2",frame=F,axes=F)
  add_letter(figlabels[1]); figlabels <- figlabels[-1]

  for(i in 1:nrow(module_means_per_dataset[[j]])){

    xx <- module_means_per_dataset[[j]][i,][sample_use]
    xx <- xx / max(xx)
    o <- order(xx)
    plot(UMAP_TRX[o,],
         col=colorRampPalette( c("grey95","navy")) (99)[ 1 + (xx[o]*98) ],
         main=rownames(module_means_per_dataset[[j]])[i],
         pch=16,xlab="UMAP1",ylab="UMAP2",frame=F,axes=F,font.main=1,cex.main=1)
  }
}
```
